---
title: "Optimizing Protein Yield"
subtitle: "RSM with BO"
author: "Jose Manuel Rivera Espejo"
date: last-modified
date-format: "DD MMMM YYYY"
bibliography: references.bib
title-slide-attributes: 
  data-notes: |
    Thank you for having me, Lynn and Francisco
include-in-header:
  - text: |
      <style>
      #title-slide .title {
        font-size: 1.5em;
      }
      </style>
---

## The contents {style="font-size:60%;"}

1. Motivating example
2. Step 1: Identify relevant factors 
3. Step 2: Steepest ascend
4. Step 3: Bayesian Optimization
5. The reality 
6. FAIR principles in practice

```{r}
#| echo: false
#| output: false
#| code-fold: true
#| code-summary: "code"

# working environment ####

# cleaning R start
rm(list=ls()); gc()

# loading libraries
libraries = c( 
  # general purpose
  'here',
  # experimental design
  'FrF2','daewr','rsm'
) 
# sapply(libraries, install.packages, character.only=T)
sapply(libraries, require, character.only=T)

# sourcing used defined functions (udf)
source( file.path(here(), 'code', 'udf.R') )

seed = 12345

```


---

## Motivating example {style="font-size:60%;"}

**The context:**

- We have a high-value protein production process,
- We known that seven ($7$) factors may potentially affect the process yield ($Y$),
- We certain that our current operating conditions are "sub-optimal"

| Variable | Factor           | Range       | Units |
| :---     | :---             | :---        | :---  |
| $A$      | pH               | $0.0-14.0$  |       |
| $B$      | Temp             | $0.0-100.0$ | C     |
| $C$      | Incubation time  | $0-96$      | h     |
| $D$      | Agitation speed  | $0-1000$    | RPM   |
| $E$      | Initial glucose  | $0-30$      | g/l   |
| $F$      | Dissolved oxygen | $0-100$     | %     |
| $G$      | Inoculum volume  | $0-100$     | %     |

---

## Motivating example {style="font-size:60%;"}

**The conflict:** laboratory experiments are

- Expensive
- Time-consuming
- Resource-constrained

::: {.fragment}
**The goal:** 

- Identify relevant factors that affect the process (*screen*)
- Find the "peak yield" (*optimize*)
- **Restriction:** use as few runs as possible
:::


---

## Motivating example {style="font-size:60%;"}

Our current operating conditions are in a small parameter space, but there is so much **parameter space** to explore!

```{r}
#| echo: false
#| output: true
#| code-fold: true
#| code-summary: "code"

## Current operation conditions ####
k = 7 # number of factors
d0 = data.frame( matrix( c(2, 10, 5, 100, 10, 20, 20), nrow=1, ncol=7) )
names(d0) = LETTERS[1:k]

set.seed( seed )
d0$Y = true_protein_yield( X=as.matrix(d0) )

```

```{r}
#| echo: false
#| output: true
#| code-fold: true
#| code-summary: "code"

# plot
twi = list( c(1,2), c(3,2),
            c(1,3), c(3,4) )
mr = c(14, 100, 24*4, 1000, 30, 100, 100)

par( mfrow=c(2,2) )
for( j in 1:length(twi) ){
  idx = !names(d0) %in% c( names(d0)[ twi[[j]] ], 'Y' )
  nam = paste( 
    names(d0)[ twi[[j]][1] ],'\n',
    paste( names(d0)[idx], '=', d0[,idx], collapse=', ') )
  
  plot( d0[,twi[[j]]], 
        pch=19, col=rgb(0,0,0,0.5),
        ylim=c( 0, mr[ twi[[j]][2] ] ),
        xlim=c( 0, mr[ twi[[j]][1] ] ), 
        ylab=names(d0)[ twi[[j]][2] ],
        xlab=nam )
  text( d0[,twi[[j]]] + 1, labels =round(d0$Y,1) )
}
par( mfrow=c(1,1) )

```


---

## Motivating example {style="font-size:60%;"}

To facilitate the explanation, we focus on factors A and B; however, the ideas presented here apply equally to the remaining factors.

```{r}
#| echo: false
#| output: true
#| code-fold: true
#| code-summary: "code"

# plot
par( mfrow=c(1,1) )
idx = !names(d0) %in% c( names(d0)[ 1:2 ], 'Y' )
nam = paste( 
  names(d0)[ 1 ],'\n',
  paste( names(d0)[idx], '=', d0[,idx], collapse=', ') )

plot( d0[,1:2], pch=19, col=rgb(0,0,0,0.5),
      ylim=c( 0, mr[ 2 ] ), xlim=c( 0, mr[ 1 ] ), 
      ylab=names(d0)[ 2 ], xlab=nam )
text( d0[,1:2] + 0.5, labels =round(d0$Y,1) )

```


---

## Motivating example {style="font-size:60%;"}

Moreover, we focus on a restricted region within the seven ($7$) factor space, illustrated here using factors A and B.

```{r}
#| echo: false
#| output: true
#| code-fold: true
#| code-summary: "code"

# plot
par( mfrow=c(1,1) )
idx = !names(d0) %in% c( names(d0)[ 1:2 ], 'Y' )
nam = paste( 
  names(d0)[ 1 ],'\n',
  paste( names(d0)[idx], '=', d0[,idx], collapse=', ') )

plot( d0[,1:2], pch=19, col=rgb(0,0,0,0.5),
      ylim=c( 0, mr[ 2 ] ), xlim=c( 0, mr[ 1 ] ), 
      ylab=names(d0)[ 2 ], xlab=nam )
rect( 0, 0, 4, 20, lty=2 )
text( d0[,1:2] + 0.5, labels =round(d0$Y,1) )

```

---

## Motivating example {style="font-size:60%;"}

Moreover, we focus on a restricted region within the seven ($7$) factor space, illustrated here using factors A and B.

```{r}
#| echo: false
#| output: true
#| code-fold: true
#| code-summary: "code"

# plot
par( mfrow=c(1,1) )
idx = !names(d0) %in% c( names(d0)[ 1:2 ], 'Y' )
nam = paste( 
  names(d0)[ 1 ],'\n',
  paste( names(d0)[idx], '=', d0[,idx], collapse=', ') )

plot( d0[,1:2], pch=19, col=rgb(0,0,0,0.5),
      ylim=c( 0, 20 ), xlim=c( 0, 4 ), 
      ylab=names(d0)[ 2 ], xlab=nam )
text( d0[,1:2] + 0.15, labels =round(d0$Y,1) )

```

---

## Step 1: Identify relevant factors {style="font-size:60%;"}

Two options: 

- A **$2^{k}$ full factorial (CRFD)** design [@Lawson_2015] with $k=7$ factors would take $2^{k}=2^{7}=128$ experimental runs.

- A **resolution III fractional factorial (CRFF)** design [@Lawson_2015] with $k=7$ factors and $p=4$ generators would take $2^{k-p} = 2^{7-4} = 8$ experimental runs.

Benefits:

- A **$2^{k}$ CRFD** design allows you to estimate an intercept, all $7$ main effects, all $21$ two-way interactions, all $7$ pure quadratic terms, and more.

- A **resolution III CRFF** design only allows you estimate an intercept, and all $7$ main effects

:::{.notes}
- CAREFUL resolution III CRFF has main effects aliased with some 2fi
:::

---

## Step 1: Identify relevant factors {style="font-size:60%;"}

A **resolution III CRFF** design [@Lawson_2015] with $k=7$ factors and $p=4$ generators, have the following characteristics:

```{r}
#| echo: false
#| output: true
#| code-fold: true
#| code-summary: "code"

# design CRFF
set.seed( seed )
dsg = FrF2( nfactors=k, resolution=3 )

# checking design
colormap(dsg, mod=2)

```

:::{.notes}
- resolution III CRFF: main effects are aliased with 2fi

```{r}
#| echo: false
#| output: true
#| code-fold: true
#| code-summary: "code"

# checking design
print('Aliases')
y = runif( nrow(dsg), 0, 1)
aliases( lm( y~(.)^2, data=dsg) )

```
:::


---

## Step 1: Identify relevant factors {style="font-size:60%;"}

How a **resolution III CRFF** looks like?

```{r}
#| echo: false
#| output: true
#| code-fold: true
#| code-summary: "code"

# convert to data.frame
d = data.frame( as.matrix(dsg[,1:k]) ) # design
d = sapply(d, as.integer)
d = data.frame(d)

# create natural variables
nv = rbind( 
  d0[,-ncol(d0)], # central points
  c(3-1, 15-5, 6-4, 150-50, 15-5, 30-10, 30-10) # ranges to test
)
for( j in 1:k ){
  d[,j] = d[,j]*nv[2,j]/2 + nv[1,j] 
}

```

```{r}
#| echo: false
#| output: true
#| code-fold: true
#| code-summary: "code"

# plot
par( mfrow=c(1,1) )
idx = !names(d0) %in% c( names(d0)[ 1:2 ], 'Y' )
nam = paste( 
  names(d0)[ 1 ],'\n',
  paste( names(d0)[idx], '=', d0[,idx], collapse=', ') )

plot( d0[,1:2], pch=19, col=rgb(0,0,0,0.5),
      ylim=c( 0, 20 ), xlim=c( 0, 4 ), 
      ylab=names(d0)[ 2 ], xlab=nam )
points( d[,1:2], pch=19, col=rgb(0,0,1,0.5) )
legend( 'topleft', legend=c( 'CRFF design','current OC' ),
        fill=c(rgb(0,0,1,0.5), rgb(0,0,0,0.5)), bty='n' )


```

::: {.notes}
This design happen at every combination of the $7$ factors
:::


---

## Step 1: Identify relevant factors {style="font-size:60%;"}

After conducting the experiments, we test which factors are relevant:

```{r}
#| echo: false
#| output: true
#| code-fold: true
#| code-summary: "code"

# measure
set.seed( seed )
d$Y = true_protein_yield( X = as.matrix( d[,1:k] ) )

# model for all First Order (FO) effects
model1 = lm( Y ~ A + B + C + D + E + F + G, data=d)
summary(model1)

```


---

## Step 1: Identify relevant factors {style="font-size:60%;"}

Thanks to the Central Limit Theorem (CLT), we can still identify significant effects.

```{r}
#| echo: false
#| output: true
#| code-fold: true
#| code-summary: "code"

# model for all First Order (FO) effects
fullnormal( coef(model1), alpha=0.025 )

```

::: {.notes}
We can use this plot thanks to CLT

NO relevance of $D$, which could indicate the presence of 2fi between $A$ and $B$

```{r}
#| echo: false
#| output: true
#| code-fold: true
#| code-summary: "code"

# checking design
print('Aliases')
y = runif( nrow(dsg), 0, 1)
aliases( lm( y~(.)^2, data=dsg) )

```

:::


---

## Step 1: Identify relevant factors {style="font-size:60%;"}

After conducting the experiments, we test which factors are relevant:

```{r}
#| echo: false
#| output: true
#| code-fold: true
#| code-summary: "code"

# reduced model for First Order (FO) effects
model2 = lm( Y ~ A + B + C, data=d)
summary(model2)

```


---

## Step 1: Identify relevant factors {style="font-size:60%;"}

After conducting the experiments, we test which factors are relevant:

```{r}
#| echo: false
#| output: true
#| code-fold: true
#| code-summary: "code"

# reduced model for First Order (FO) effects
model3 = lm( Y ~ A + B, data=d)
summary(model3)

```


---

## Step 2: Steepest ascend {style="font-size:60%;"}

In order to use methods from **Response Surface Methodology (RMS)** [@Lawson_2015] we need to add some center points and build a model that describes the yield space:

```{r}
#| echo: false
#| output: true
#| code-fold: true
#| code-summary: "code"

# experiment block
d$block = 'screen'

# add some center points
for( i in 1:4 ){
  d = rbind( d, data.frame( d0[,-ncol(d0)], Y=NA, block='center') )
}

# plot
par( mfrow=c(1,1) )
idx = !names(d0) %in% c( names(d0)[ 1:2 ], 'Y' )
nam = paste( 
  names(d0)[ 1 ],'\n',
  paste( names(d0)[idx], '=', d0[,idx], collapse=', ') )

plot( d[,1:2], pch=19, col=rgb(0,0,1,0.5),
      ylim=c( 0, 20 ), xlim=c( 0, 4 ), 
      ylab=names(d0)[ 2 ], xlab=nam )
legend( 'topleft', legend=c( 'CRFF design +\ncenter points' ),
        fill=rgb(0,0,1,0.5), bty='n' )

```

---

## Step 2: Steepest ascend {style="font-size:60%;"}

After collecting the outcome resulting from the experimental design we see:

```{r}
#| echo: false
#| output: true
#| code-fold: true
#| code-summary: "code"

# measure
set.seed( seed )
d$Y[is.na(d$Y)] = true_protein_yield( X = as.matrix( d[is.na(d$Y),1:k] ) )

# model
model_rsm = rsm( Y ~ FO(A,B), data=d )
summary(model_rsm)

```

```{r}
#| echo: false
#| output: true
#| code-fold: true
#| code-summary: "code"

# direction
sa = steepest(model_rsm, dist=1, descent=F)

# new point
new = d0
check = T
set.seed( seed )
while( check ){
  new$A = new$A + 1
  if( new$A > mr[1] ){
    break
  }
  new$B = new$B + with(sa, A/B)*(new$A-2)
  new$Y = true_protein_yield( X = as.matrix( new[,-ncol(new)] ) )
  d = rbind( d, data.frame(new, block='ascend') )
  check = with(d, Y[nrow(d)] - Y[nrow(d)-1] > 0 )
}

```


---

## Step 2: Steepest ascend {style="font-size:60%;"}

How does the space look like?

```{r}
#| echo: false
#| output: true
#| code-fold: true
#| code-summary: "code"

# plot
par( mfrow=c(1,1) )
idx = !names(d0) %in% c( names(d0)[ 1:2 ], 'Y' )
nam = paste( 
  names(d0)[ 1 ],'\n',
  paste( names(d0)[idx], '=', d0[,idx], collapse=', ') )

contour( model_rsm, ~ A+B, image=T,
         ylim=c( 0, 20 ),
         xlim=c( 0, 4 ) )
points( d[ d$block!='ascend', 1:2 ], 
        pch=19, col=rgb(0,0,0,0.5) )

# equation of a line
point1 = d[which( d$block=='center')[1], 1:2 ] 
point2 = d[which( d$block=='ascend')[1], 1:2 ] 
b = ( point2[,2] - point1[,2] ) / ( point2[,1] - point1[,1] )
a = point1[,2] - b*point1[,1]
abline( a=a, b=b, lty=2 )

```

---

## Step 2: Steepest ascend {style="font-size:60%;"}

Now we proceed to ascend towards a maximum.

```{r}
#| echo: false
#| output: true
#| code-fold: true
#| code-summary: "code"

# plot
par( mfrow=c(1,1) )
idx = !names(d0) %in% c( names(d0)[ 1:2 ], 'Y' )
nam = paste( 
  names(d0)[ 1 ],'\n',
  paste( names(d0)[idx], '=', d0[,idx], collapse=', ') )

contour( model_rsm, ~ A+B, image=T,
         ylim=c( 0, 100 ),
         xlim=c( 0, 14 ) )
points( d[, 1:2 ], 
        pch=19, col=rgb(0,0,0,0.5) )
text( d[ d$block=='ascend', 1:2 ] - 0.5, 
      labels = round( d$Y[ d$block=='ascend'],1) )

# equation of a line
abline( a=a, b=b, lty=2 )

```


---

## Step 2: Steepest ascend {style="font-size:60%;"}

And we focus our attention on the highest yield area:

```{r}
#| echo: false
#| output: true
#| code-fold: true
#| code-summary: "code"

# plot
par( mfrow=c(1,1) )
idx = !names(d0) %in% c( names(d0)[ 1:2 ], 'Y' )
nam = paste( 
  names(d0)[ 1 ],'\n',
  paste( names(d0)[idx], '=', d0[,idx], collapse=', ') )

contour( model_rsm, ~ A+B, image=T,
         ylim=c( 0, 100 ),
         xlim=c( 0, 14 ) )
points( d[, 1:2 ], 
        pch=19, col=rgb(0,0,0,0.5) )
text( d[ d$block=='ascend', 1:2 ] - 0.5, 
      labels = round( d$Y[ d$block=='ascend'],1) )

# equation of a line
abline( a=a, b=b, lty=2 )
rect( 11, 60, 14, 100, lty=2 )

```


---

## Step 2: Steepest ascend {style="font-size:60%;"}

And we focus our attention on the highest yield area:

```{r}
#| echo: false
#| output: true
#| code-fold: true
#| code-summary: "code"

# plot
par( mfrow=c(1,1) )
idx = !names(d0) %in% c( names(d0)[ 1:2 ], 'Y' )
nam = paste( 
  names(d0)[ 1 ],'\n',
  paste( names(d0)[idx], '=', d0[,idx], collapse=', ') )

contour( model_rsm, ~ A+B, image=T,
         ylim=c( 60, 100 ),
         xlim=c( 11, 14 ) )
points( d[, 1:2 ], 
        pch=19, col=rgb(0,0,0,0.5) )
text( d[ d$block=='ascend', 1:2 ] - 0.1, 
      labels = round( d$Y[ d$block=='ascend'],1) )

```

---

## Step 3: Bayesian Optimization {style="font-size:60%;"}

Within this high-yield region, we implement **Bayesian Optimization (BO)** [@Sweet_2023].

Specifically:

- We use relevant **initial training data** from the steepest ascent procedure
  
    In this case, the three design points located within the high-yield region.
    
- We use a **surrogate model** to approximate the yield surface for a set of **candidate data points**

    Here, a Gaussian Process with Radial Basis Function (RBF) kernel [@Rasmussen_et_al_2015; @Nielsen_2023] is employed .

- We use an **acquisition function** to determine the next point to evaluate in the search space

    In this case, we use the Expected Improvement function (see illustration).

- We define a **stopping criterion** to determine when to terminate the search

    Because a central composite design would require $2^2 = 4$ factorial points, $2^2 = 4$ axial (face) points, and $4$ center points, for a total of $12$ experimental runs [@Lawson_2015], the optimization procedure will use $8$ runs.


---

## Step 3: Bayesian Optimization {style="font-size:60%;"}

How does the training and candidate data look like?

```{r}
#| echo: false
#| output: true
#| code-fold: true
#| code-summary: "code"

# initial training points
dBO = tail(d, 3)
x_train = dBO[, c('A','B') ]
y_train = dBO$Y

# prediction points
x_predict = data.frame( 
  expand.grid( A=seq(11,14,by=0.2), B=seq(60,100,by=5)) )

```

```{r}
#| echo: false
#| output: true
#| code-fold: true
#| code-summary: "code"

# plot
par( mfrow=c(1,1) )
idx = !names(d0) %in% c( names(d0)[ 1:2 ], 'Y' )
nam = paste( 
  names(d0)[ 1 ],'\n',
  paste( names(d0)[idx], '=', d0[,idx], collapse=', ') )

plot( d[, 1:2 ], 
        pch=19, col=rgb(0,0,0,0.5),
        ylim=c( 59, 115 ),
        xlim=c( 10.95, 14.05 ) )
text( d[ d$block=='ascend', 1:2 ] - 0.1, 
      labels = round( d$Y[ d$block=='ascend'],1) )
points( x_predict, pch=19, col=rgb(0,0,1,0.5) )
legend( 'topleft', legend=c( 'Training data', 'Candidate points' ),
        fill=c( rgb(0,0,0,0.5), rgb(0,0,1,0.5) ), bty='n' )

```


---

## Step 3: Bayesian Optimization {style="font-size:60%;"}

TO DO


---

## The reality {style="font-size:60%;"}

The "unknown" **true model** for protein yield ($Y$) was:

$$
Y = 85 - 0.4(A-12)^2 - 0.02(B-70)^2 + \sum_{j=3}^{7} 0.01 \cdot X_{j} + 0.1(A-12)(B-70) + e; \quad \; e \sim N(0, 2)
$$

"unknown" **true** protein yield surface:

```{r}
#| echo: false
#| output: true
#| code-fold: true
#| code-summary: "code"

# setting seed for plot replication
set.seed( seed )

# simulation
X = true_covariates( n=1000 )
Y = true_protein_yield( X=X )
dT = data.frame( X, Y)
m = rsm( Y ~ FO(X1,X2) + TWI(X1,X2) + PQ(X1,X2), data=dT )

# plot
contour( m, ~ X1+X2, image=T,
         xlabs=c("pH", "Temperature (ºC)") )
abline( v=12, h=70, lty=2, col=rgb(0,0,0,0.3) )

```

::: {.notes}
- In reality the **true model** is not available to us.
- However, simulation is useful to showcase the potential of the experimental strategies.
:::


---

## The reality {style="font-size:60%;"}

The "unknown" **true model** for protein yield ($Y$) was:

$$
Y = 85 - 0.4(A-12)^2 - 0.02(B-70)^2 + \sum_{j=3}^{7} 0.01 \cdot X_{j} + 0.1(A-12)(B-70) + e; \quad \; e \sim N(0, 2)
$$

"unknown" **true** protein yield surface with screen, ascend, and optimization points:

```{r}
#| echo: false
#| output: true
#| code-fold: true
#| code-summary: "code"

# plot
contour( m, ~ X1+X2, image=T,
         xlabs=c("pH", "Temperature (ºC)") )
abline( v=12, h=70, lty=2, col=rgb(0,0,0,0.3) )
points( d[ d$block!='ascend', 1:2 ], pch=19, col=rgb(0,0,0,0.5),
        ylim=c( 59, 115 ), xlim=c( 10.95, 14.05 ) )
points( d[ d$block=='ascend', 1:2 ], pch=19, col=rgb(0,0,1,0.5),
      ylim=c( 59, 115 ), xlim=c( 10.95, 14.05 ) )
# points( d[ d$block=='BO', 1:2 ], pch=19, col=rgb(1,0,0,0.5),
#         ylim=c( 59, 115 ), xlim=c( 10.95, 14.05 ) )
legend( 'left', legend=c( 'Screen', 'Ascend', 'Optimize' ),
        fill=c( rgb(0,0,0,0.5), rgb(0,0,1,0.5), rgb(1,0,0,0.5) ), bty='n' )


```

---

## FAIR principles in practice {style="font-size:60%;"}

This presentation is built to be a **research object**, not just a slide deck.

- **Findable (F):** Searchable text and structured metadata ensure the content is discoverable. 
    
    Next step: Provision of a DOI via Zenodo or OSF.

- **Accesible (A):** Publicly hosted on GitHub. 

    Anyone with an internet connection can retrieve the slides and underlying data.

- **Interoperable (I):** Built using Open Source (Quarto/`R`).

    Seamlessly integrates into any analysis workflow and automated pipelines.
    
- **Reusable (R):** CC-BY 4.0 Licensed.

    Includes all `R` dependencies and source code to fully replicate every simulation and analysis. 


---

## Thank you {.center-xy}

::: {.notes}
To conclude I can say that, 

:::


---

## Questions? {.center-xy}

---

## Annex: additional considerations {style="font-size:60%;"}

In a real-world setting. the identification of those $7$ factors probably comes from experiments that seek to determine **Sources of Variability**. e.g.,

* **RSE (Random Statistical Error):** Natural fluctuations in biological systems.
* **NSE (Non-Systematic Error):** Potential issues with equipment or batch effects.

---

## Annex: additional considerations {style="font-size:60%;"}

What happens when the problem grows?

* **Multiple Objectives:** If we need high **Yield** but low **Waste**, we employ **Desirability Functions** to find the "Sweet Spot" that balances both.
* **Extrapolation:**
    * Empirical models (RSM/BO) are dangerous outside their tested range.
    * If we must predict outside our data, we shift toward **Theoretical/Mechanistic Models** based on biological first principles.


## References {style="font-size:60%;"}

:::{#refs style="font-size:80%;"}

:::
