---
title: "Optimizing Protein Yield"
subtitle: "A comparative approach: RSM vs. BO"
author: "Jose Manuel Rivera Espejo"
date: last-modified
date-format: "DD MMMM YYYY"
bibliography: references.bib
title-slide-attributes: 
  data-notes: |
    Thank you for having me, Lynn and Francisco
include-in-header:
  - text: |
      <style>
      #title-slide .title {
        font-size: 1.5em;
      }
      </style>
---

## The contents {style="font-size:60%;"}

1. Motivating example
2. 
3. 
4. 
5.  
6. 
7. 
8. 

```{r}
#| echo: false
#| output: false
#| code-fold: true
#| code-summary: "code"

# working environment ####

# cleaning R start
rm(list=ls()); gc()

# loading libraries
libraries = c( 
  # general purpose
  'here',
  # experimental design
  'FrF2','daewr','rsm'
) 
# sapply(libraries, install.packages, character.only=T)
sapply(libraries, require, character.only=T)

# sourcing used defined functions (udf)
source( file.path(here(), 'code', 'udf.R') )

seed = 12345

```


---

## Motivating example {style="font-size:60%;"}

::: {.fragment}
**The context:**

::: incremental
- We have a high-value protein production process,
- We known that seven ($7$) factors may potentially affect the process yield ($Y$),
- We are uncertain if our current operating conditions are "sub-optimal"
:::
:::

::: {.fragment}
**The conflict:** laboratory experiments are

::: incremental
- Expensive
- Time-consuming
- Resource-constrained
:::
:::

::: {.fragment}
**The goal:** 

::: incremental
- Identify relevant factors that affect the process (*screen*)
- Find the "peak yield" (*optimize*)
- **Restriction:** use as few runs as possible
:::
:::


---

## Step 1: Identify relevant factors {style="font-size:60%;"}

Our current operating conditions are in a small space, but there is so much **parameter space** to explore!

```{r}
#| echo: false
#| output: true
#| code-fold: true
#| code-summary: "code"

## Current operation conditions ####
k = 7 # number of factors
d0 = data.frame( matrix( c(2, 10, 5, 100, 10, 20, 20), nrow=1, ncol=7) )
names(d0) = LETTERS[1:k]

set.seed( seed )
d0$Y = true_protein_yield( X=as.matrix(d0) )

```

```{r}
#| echo: false
#| output: true
#| code-fold: true
#| code-summary: "code"

# plot
twi = list( c(1,2), c(3,2),
            c(1,3), c(3,4) )
mr = c(14, 100, 24*4, 1000, 30, 100, 100)

par( mfrow=c(2,2) )
for( j in 1:length(twi) ){
  idx = !names(d0) %in% c( names(d0)[ twi[[j]] ], 'Y' )
  nam = paste( 
    names(d0)[ twi[[j]][1] ],'\n',
    paste( names(d0)[idx], '=', d0[,idx], collapse=', ') )
  
  plot( d0[,twi[[j]]], 
        pch=19, col=rgb(0,0,0,0.5),
        ylim=c( 0, mr[ twi[[j]][2] ] ),
        xlim=c( 0, mr[ twi[[j]][1] ] ), 
        ylab=names(d0)[ twi[[j]][2] ],
        xlab=nam )
  text( d0[,twi[[j]]] + 1, labels =round(d0$Y,1) )
}
par( mfrow=c(1,1) )

```


---

## Step 1: Identify relevant factors {style="font-size:60%;"}

Our current operating conditions are in a small space, but there is so much **parameter space** to explore!

```{r}
#| echo: false
#| output: true
#| code-fold: true
#| code-summary: "code"

# plot
twi = list( c(1,2) )

par( mfrow=c(1,1) )
idx = !names(d0) %in% c( names(d0)[ twi[[1]] ], 'Y' )
nam = paste( 
  names(d0)[ twi[[1]][1] ],'\n',
  paste( names(d0)[idx], '=', d0[,idx], collapse=', ') )

plot( d0[,twi[[1]]], 
      pch=19, col=rgb(0,0,0,0.5),
      ylim=c( 0, mr[ twi[[1]][2] ] ),
      xlim=c( 0, mr[ twi[[1]][1] ] ), 
      ylab=names(d0)[ twi[[1]][2] ],
      xlab=nam )
text( d0[,twi[[1]]] + 0.5, labels =round(d0$Y,1) )

```


---

## Step 1: Identify relevant factors {style="font-size:60%;"}

Our current operating conditions are in a small space, but there is so much **parameter space** to explore!

```{r}
#| echo: false
#| output: true
#| code-fold: true
#| code-summary: "code"

# plot
twi = list( c(1,2) )

par( mfrow=c(1,1) )
idx = !names(d0) %in% c( names(d0)[ twi[[1]] ], 'Y' )
nam = paste( 
  names(d0)[ twi[[1]][1] ],'\n',
  paste( names(d0)[idx], '=', d0[,idx], collapse=', ') )

plot( d0[,twi[[1]]], 
      pch=19, col=rgb(0,0,0,0.5),
      ylim=c( 0, mr[ twi[[1]][2] ] ),
      xlim=c( 0, mr[ twi[[1]][1] ] ), 
      ylab=names(d0)[ twi[[1]][2] ],
      xlab=nam )
rect( 0, 0, 4, 20)
text( d0[,twi[[1]]] + 0.5, labels =round(d0$Y,1) )

```

---

## Step 1: Identify relevant factors {style="font-size:60%;"}

Our current operating conditions are in a small space, but there is so much **parameter space** to explore!

```{r}
#| echo: false
#| output: true
#| code-fold: true
#| code-summary: "code"

# plot
twi = list( c(1,2) )

par( mfrow=c(1,1) )
idx = !names(d0) %in% c( names(d0)[ twi[[1]] ], 'Y' )
nam = paste( 
  names(d0)[ twi[[1]][1] ],'\n',
  paste( names(d0)[idx], '=', d0[,idx], collapse=', ') )

plot( d0[,twi[[1]]], 
      pch=19, col=rgb(0,0,0,0.5),
      ylim=c( 0, 20 ),
      xlim=c( 0, 4 ), 
      ylab=names(d0)[ twi[[1]][2] ],
      xlab=nam )
text( d0[,twi[[1]]] + 0.15, labels =round(d0$Y,1) )

```

---

## Step 1: Identify relevant factors {style="font-size:60%;"}

Which of our $7$ factors affect the process yield?

| Variable | Factor           | Range       | Units | Role         |
| :---     | :---             | :---        | :---  | :---         |
| $A$      | pH               | $0.0-14.0$  |       | Relevant     |
| $B$      | Temp             | $0.0-100.0$ | C     | Relevant     |
| $C$      | Incubation time  | $0-96$      | h     | Non-relevant |
| $D$      | Agitation speed  | $0-1000$    | RPM   | Non-relevant |
| $E$      | Initial glucose  | $0-30$      | g/l   | Non-relevant |
| $F$      | Dissolved oxygen | $0-100$     | %     | Non-relevant |
| $G$      | Inoculum volume  | $0-100$     | %     | Non-relevant |

::: {.fragment}
Two options: 

::: incremental
- A **full factorial (CRFD)** design with $k=7$ factors would take $2^{k}=2^{7}=128$ experimental runs.

- A **resolution III fractional factorial (CRFF)** design with $k=7$ factors would take $2^{k-p} = 2^{7-4} = 8$ experimental runs.
:::
:::

:::{.notes}
- you have at least $7$ main effects and $8$ runs (only $1$ df)
- resolution III CRFF: $p=4$ main effects are aliased with 2fi
:::

---

## Step 1: Identify relevant factors {style="font-size:60%;"}

A **resolution III fractional factorial (CRFF)** design with $k=7$ factors, have the following characteristics:

```{r}
#| echo: false
#| output: true
#| code-fold: true
#| code-summary: "code"

# design CRFF
set.seed( seed )
dsg = FrF2( nfactors=k, resolution=3 )

# checking design
colormap(dsg, mod=2)

```

:::{.notes}
- resolution III CRFF: $p=4$ main effects are aliased with 2fi

```{r}
#| echo: false
#| output: true
#| code-fold: true
#| code-summary: "code"

# checking design
print('Aliases')
y = runif( nrow(dsg), 0, 1)
aliases( lm( y~(.)^2, data=dsg) )

```
:::


---

## Step 1: Identify relevant factors {style="font-size:60%;"}

How a **resolution III CRFF** looks like?

```{r}
#| echo: false
#| output: true
#| code-fold: true
#| code-summary: "code"

# convert to data.frame
d = data.frame( as.matrix(dsg[,1:k]) ) # design
d = sapply(d, as.integer)
d = data.frame(d)

# create natural variables
nv = rbind( 
  d0[,-ncol(d0)], # central points
  c(3-1, 15-5, 6-4, 150-50, 15-5, 30-10, 30-10) # ranges to test
)
for( j in 1:k ){
  d[,j] = d[,j]*nv[2,j]/2 + nv[1,j] 
}

```

```{r}
#| echo: false
#| output: true
#| code-fold: true
#| code-summary: "code"

# plot
twi = list( c(1,2) )

par( mfrow=c(1,1) )
idx = !names(d0) %in% c( names(d0)[ twi[[1]] ], 'Y' )
nam = paste( 
  names(d0)[ twi[[1]][1] ],'\n',
  paste( names(d0)[idx], '=', d0[,idx], collapse=', ') )

plot( d0[,twi[[1]]], 
      pch=19, col=rgb(0,0,0,0.5),
      ylim=c( 0, 20 ),
      xlim=c( 0, 4 ), 
      ylab=names(d0)[ twi[[1]][2] ],
      xlab=nam )
points( d[,1:2], pch=19, col=rgb(0,0,1,0.5) )

```


---

## Step 1: Identify relevant factors {style="font-size:60%;"}

After conducting the experiments, we test which factors are relevant:

```{r}
#| echo: false
#| output: true
#| code-fold: true
#| code-summary: "code"

# measure
set.seed( seed )
d$Y = true_protein_yield( X = as.matrix( d[,1:k] ) )

# model for all First Order (FO) effects
model1 = lm( Y ~ A + B + C + D + E + F + G, data=d)
summary(model1)

```


---

## Step 1: Identify relevant factors {style="font-size:60%;"}

After conducting the experiments, we test which factors are relevant:

```{r}
#| echo: false
#| output: true
#| code-fold: true
#| code-summary: "code"

# model for all First Order (FO) effects
halfnorm( coef(model1) )

```

---

## Step 1: Identify relevant factors {style="font-size:60%;"}

After conducting the experiments, we test which factors are relevant:

```{r}
#| echo: false
#| output: true
#| code-fold: true
#| code-summary: "code"

# reduced model for First Order (FO) effects
model2 = lm( Y ~ A + B, data=d)
summary(model2)

```


---

## Step 2: Steepest ascend {style="font-size:60%;"}

```{r}
#| echo: false
#| output: true
#| code-fold: true
#| code-summary: "code"

# experiment block
d$block = 'screen'

# add some center points
for( i in 1:4 ){
  d = rbind( d, data.frame( d0[,-ncol(d0)], Y=NA, block='center') )
}

set.seed( seed )
d$Y[is.na(d$Y)] = true_protein_yield( X = as.matrix( d[is.na(d$Y),1:k] ) )

# model
model_rsm = rsm( Y ~ FO(A,B), data=d )
summary(model_rsm)

# direction
sa = steepest(model_rsm, dist=1, descent=F)

# new point
new = d0
check = T
set.seed( seed )
while( check ){
  new$A = new$A + 1
  new$B = new$B + with(sa, A/B)*(new$A-2)
  new$Y = true_protein_yield( X = as.matrix( new[,-ncol(new)] ) )
  d = rbind( d, data.frame(new, block='ascend') )
  check = with(d, Y[nrow(d)] - Y[nrow(d)-1] > 0 )
}

```


---

## Step 2: Steepest ascend {style="font-size:60%;"}

```{r}
#| echo: false
#| output: true
#| code-fold: true
#| code-summary: "code"

# plot
twi = list( c(1,2) )

par( mfrow=c(1,1) )
idx = !names(d0) %in% c( names(d0)[ twi[[1]] ], 'Y' )
nam = paste( 
  names(d0)[ twi[[1]][1] ],'\n',
  paste( names(d0)[idx], '=', d0[,idx], collapse=', ') )

contour( model_rsm, ~ A+B, image=T,
         ylim=c( 0, 20 ),
         xlim=c( 0, 4 ) )
points( d[ d$block!='ascend', twi[[1]] ], 
        pch=19, col=rgb(0,0,0,0.5) )

# equation of a line
point1 = d[which( d$block=='center')[1], twi[[1]] ] 
point2 = d[which( d$block=='ascend')[1], twi[[1]] ] 
b = ( point2[,2] - point1[,2] ) / ( point2[,1] - point1[,1] )
a = point1[,2] - b*point1[,1]
abline( a=a, b=b, lty=2 )

```

---

## Step 2: Steepest ascend {style="font-size:60%;"}

```{r}
#| echo: false
#| output: true
#| code-fold: true
#| code-summary: "code"

# plot
twi = list( c(1,2) )

par( mfrow=c(1,1) )
idx = !names(d0) %in% c( names(d0)[ twi[[1]] ], 'Y' )
nam = paste( 
  names(d0)[ twi[[1]][1] ],'\n',
  paste( names(d0)[idx], '=', d0[,idx], collapse=', ') )

contour( model_rsm, ~ A+B, image=T,
         ylim=c( 0, 100 ),
         xlim=c( 0, 14 ) )
points( d[, twi[[1]] ], 
        pch=19, col=rgb(0,0,0,0.5) )
text( d[ d$block=='ascend', twi[[1]] ] - 0.5, 
      labels = round( d$Y[ d$block=='ascend'],1) )

# equation of a line
abline( a=a, b=b, lty=2 )

```

---

## Motivating example {style="font-size:60%;"}

We have an "unknown" **true model** that relates seven ($7$) variables to the yield of a protein ($Y$):

$$
Y = \text{inv_logit} \left[ \gamma_{0} - \gamma_{1}(pH - c_{1})^2 - \gamma_{2}(T - c_{2})^2 + \sum_{j=3}^{7} \gamma_{j} \cdot X_{j} + \gamma_{8}(pH - c_{1})(T - c_{2}) + e \right]
$$

with error $e \sim N(0,\sigma_{e})$ and,



::: {.notes}
- In reality the **true model** is not available to us.
- However, simulation is useful to showcase the potential of the experimental strategies.
:::

## Motivating example {style="font-size:60%;"}

But, how does this look like?

- The peak yield ($100\%$) is at $pH=10$ and $T=60$ 



---

## Additional Considerations: Variability {style="font-size:60%;"}

In a real-world setting at Excelra, we must acknowledge that not all noise is equal.

* **Sources of Variability:**
    * **RSE (Random Statistical Error):** Natural fluctuations in biological systems.
    * **NSE (Non-Systematic Error):** Potential issues with equipment or batch effects.
* **Design Intent:** While our current project focuses on optimization, we keep these sources in mind to ensure the process remains **robust**, not just high-yielding.

---

## Additional Considerations: Complexity {style="font-size:60%;"}

What happens when the problem grows?

* **Multiple Objectives:** If we need high **Yield** but low **Waste**, we employ **Desirability Functions** to find the "Sweet Spot" that balances both.
* **Extrapolation:**
    * Empirical models (RSM/BO) are dangerous outside their tested range.
    * If we must predict outside our data, we shift toward **Theoretical/Mechanistic Models** based on biological first principles.

---

## Thank you {.center-xy}

::: {.notes}
To conclude I can say that, 

1. I can bring my PhD-level expertise to the Leuven office.
3. I am ready to operationalize these workflows starting from the **20 of April**.
:::


---

## Questions? {.center-xy}

---

## References {style="font-size:60%;"}

:::{#refs style="font-size:80%;"}

:::
